What We Eval in the Shadows

### 2021-09-03 paper/main.tex
Update abstract
Fix grammar and typos in Section 1

\begin{abstract}
 Most dynamic languages allow users to turn text into code using
  various functions, often named \eval, with language-dependent semantics. The
  widespread use of these reflective functions hinders static analysis and
  prevents compilers from performing optimizations. This paper aims to provide a
  better sense of why programmers use \eval. Understanding why \eval is used in
  practice is key to finding ways to mitigate its negative impact. We have
  reasons to believe that reflective feature usage is language and application
  domain-specific; we focus on data science code written in R and compare our
  results to previous work that analyzed web programming in JavaScript. This
  paper studied \CranRunnableScripts scripts extracted from \CranPackages R
  packages, for a total of \packageAllcalls calls to \eval. We find that \eval
  is indeed in widespread use; R's \eval is more pervasive and arguably
  dangerous than what was previously reported for JavaScript.
\end{abstract}

@@

Most dynamic languages provide their users with a facility to transform
unstructured text into executable code and evaluate that code. We refer to this
reflective facility as \eval bowing to its origins in LISP, all the way back in
1956. \Eval has been much maligned over the years. In computing lore, it is as
close to a boogeyman as it gets. Yet, for McCarthy, \eval was simply the way to
write down the definition of LISP; he was surprised that someone coded it up and
offered it to end users~\cite{lisp}. Since then, reflective facilities have been
used to parameterize programs over code patterns that can be provided after the
program is written. The presence of such a feature in a language is a hallmark
of dynamism; it is a form of delayed binding as the behavior of any particular
call to \eval will only be known when the program is run, and that particular
call site is evaluated.

\paragraph{Trouble in Paradise.} Reflective facilities hinder
most attempts to reason about or apply meaning-preserving transformations to the
code using them. In practice, \eval causes static analysis techniques to lose so
much precision as to become pointless. For compilers, anything but the most
trivial, local optimizations are unsound after use of \eval. Furthermore, the
addition of arbitrary code --- code that could have been obtained from a network
connection --- as a program is running is a security vulnerability waiting to
happen. To illustrate these challenges, consider the interaction of a static
analysis tool with a dynamic language. A program analyzer computes an
over-approximation of the set of possible behaviors exhibited by the program
under study, a reflective facility must be represented by all behaviors that can
be expressed in the target language. \ie any legal sequence of instructions can
replace \eval. As dynamic languages are permissive, the tool must assume that
all functions in scope were redefined, \eg that \texttt{`+`} now opens a network
connection. A single occurrence of \eval causes the static analyzer to lose all
information about the program state and meaning of identifiers. This loss of
precision can sometimes be mitigated by analyzing the string
argument~\cite{moller03}, but if the string comes from outside the program, not
much can be done. A frustrated group of researchers argued giving up on
soundness and, instead, under-approximating dynamic features
(soundiness)~\cite{soundy}. In their words, ``a practical analysis, may pretend
that \eval does nothing unless it can precisely resolve its string argument at
compile time.'' Alas, assuming that \eval does not have side-effects or that
side-effects will not affect results is unduly optimistic.

@@

\paragraph{The Here and Now.} In this study, we investigate the
usage of \eval in programs written in the R language. R is a language designed
by statisticians for applications in data science~\cite{r,R96}. What makes
looking at R after JavaScript interesting is that, while both languages are
dynamic, they are quite different. While one can program in an object-oriented
style in R like in JavaScript, R is mostly a lazy functional language.
JavaScript was designed to run untrusted code in a browser, while R is used for
statistical computing on desktops. JavaScript is a general-purpose language used
by a vast community of programmers, while R is used for scientific computing by
data scientists and domain experts with, often, limited programming experience.
One can distinguish between library implementers who have programming experience
and a working knowledge of R, and end-users who are typically not expert
programmers with a cursory knowledge of the language.\footnote{We surveyed
end-users and did not find a single one aware that R is lazy.} Our goal is to
highlight the differences in usage between JavaScript and R and explain them in
terms of language features, application domain and programmer experience.
Hopefully, some of our observations generalize to other languages. Our data
and code are open source and publicly available:
\url{https://github.com/PRL-PRG/evalr-experiment}.

@@

\paragraph{Why do we Eval?} The results of our study suggest
that \eval is widely used for the implementation of the language, and in many
libraries. End-user code makes less frequent and less sophisticated use of
\eval. In many ways, \eval in R is as bad as it gets: it's varied, performs
side-effects and reaches to many environments. By large, the motivations for
\eval relate to various forms of language extensions and meta-programming. \Eval
is used where other languages would provide macros. But, the expressive power of
\eval is higher as it can reach arbitrarily far back in the call stack.

>>>>

\begin{abstract}
  Most dynamic languages allow users to turn text into code using various
  functions, often named \eval, with language-dependent semantics. The
  widespread use of these reflective functions hinders static analysis and
  prevents compilers from performing optimizations. This paper aims to provide a
  better sense of why programmers use \eval. Understanding why \eval is used in
  practice is key to finding ways to mitigate its negative impact. We have
  reasons to believe that reflective feature usage is language and application
  domain-specific; we focus on data science code written in R and compare our
  results to previous work that analyzed web programming in JavaScript. We
  analyze \packageAllcalls calls to \eval from \CranRunnableScripts scripts
  extracted from \CranPackages R packages. We find that \eval is indeed in
  widespread use; R's \eval is more pervasive and arguably dangerous than what
  was previously reported for JavaScript.
\end{abstract}

@@

Most dynamic languages provide their users with a facility to transform
unstructured text into executable code and evaluate that code. We refer to this
reflective facility as \eval bowing to its origins in LISP, all the way back in
1956. \Eval has been much maligned over the years. In computing lore, it is as
close to a boogeyman as it gets. Yet, for McCarthy, \eval was simply the way to
write down the definition of LISP; he was surprised that someone coded it up and
offered it to end-users~\cite{lisp}. Since then, reflective facilities have been
used to parameterize programs over code patterns that can be provided after the
program is written. The presence of such a feature in a language is a hallmark
of dynamism; it is a form of delayed binding as the behavior of any particular
call to \eval will only be known when the program is run, and that particular
call site is evaluated.

\paragraph{Trouble in Paradise.} Reflective facilities hinder
most attempts to reason about or apply meaning-preserving transformations to the
code using them. In practice, \eval causes static analysis techniques to lose so
much precision as to become pointless. For compilers, anything but the most
trivial, local optimizations are unsound after the use of \eval. Furthermore, the
addition of arbitrary code --- code that could have been obtained from a network
connection --- as a program is running is a security vulnerability waiting to
happen. To illustrate these challenges, consider the interaction of a static
analysis tool with a dynamic language. A program analyzer computes an
over-approximation of the set of possible behaviors exhibited by the program
under study, a reflective facility must be represented by all behaviors that can
be expressed in the target language. \ie any legal sequence of instructions can
replace \eval. As dynamic languages are permissive, the tool must assume that
all functions in scope were redefined, \eg that \texttt{`+`} now opens a network
connection. A single occurrence of \eval causes the static analyzer to lose all
information about the program state and meaning of identifiers. This loss of
precision can sometimes be mitigated by analyzing the string
argument~\cite{moller03}, but if the string comes from outside the program, not
much can be done. A frustrated group of researchers argued giving up on
soundness and, instead, under-approximating dynamic features
(soundiness)~\cite{soundy}. In their words, ``a practical analysis, may pretend
that \eval does nothing unless it can precisely resolve its string argument at
compile time.'' Alas, assuming that \eval does not have side-effects or that
side-effects will not affect results is unduly optimistic.

@@

\paragraph{The Here and Now.} In this study, we investigate the usage of \eval
in programs written in the R language. R is a language designed by statisticians
for applications in data science~\cite{r,R96}. What makes looking at R after
JavaScript interesting is that, while both languages are dynamic, they are quite
different. While one can program in an object-oriented style in R like in
JavaScript, R is mostly a lazy functional language. JavaScript was designed to
run untrusted code in a browser, while R is used for statistical computing on
desktops. JavaScript is a general-purpose language used by a vast community of
programmers, while R is used for scientific computing by data scientists and
domain experts with, often, limited programming experience. One can distinguish
between library implementers who have programming experience and a working
knowledge of R, and end-users who are typically not expert programmers with a
cursory knowledge of the language.\footnote{We surveyed end-users and did not
  find a single one aware that R is lazy.} Our goal is to highlight the
differences in usage between JavaScript and R and explain them in terms of
language features, application domain, and programmer experience. Hopefully,
some of our observations generalize to other languages. Our data and code are
open-source and publicly available at:

\begin{center}
  \url{https://doi.org/10.5281/zenodo.5415230}
\end{center}

@@

\paragraph{Why do we Eval?} The results of our study suggest
that \eval is widely used for the implementation of the language, and in many
libraries. End-user code makes less frequent and less sophisticated use of
\eval. In many ways, \eval in R is as bad as it gets: it's varied, performs
side-effects, and reaches many environments. By large, the motivations for
\eval relate to various forms of language extensions and meta-programming. \Eval
is used where other languages would provide macros. But, the expressive power of
\eval is higher as it can reach arbitrarily far back in the call stack.
